maxent model avec xvalidation (year-based test training)

### 2019-07-04 src/maxent_alldata_avec_xvalidation.Rmd
fold-aic-models

```{r remove unnecessary columns}
presback <- subset(presback, select = -c(cell_id, id, decimalLatitude, decimalLongitude, datecollected, institutioncode, individualcount, depth, resname, originalscientificname, collectioncode, day, nafo_zone, gear, longitude_meters, latitude_meters, total_cell_obs_xy, total_cell_obs_xyt, bottom_depth, XXtotal_cell_obs_xyzt, temp_celsius_depth, temp_celsius_surface, longitude_meters.1, latitude_meters.1, bottom_depth_glorys, longitude_meters.2, latitude_meters.2, cell_id_3d, total_cell_obs_xyzt, cell_id_xyzt, month, depthlayerno, depth_layer))
colnames(presback)
```

@@

```{r}
allvar <- c("occurrence", "year", "temp_surface", "temp_depth", "salinity_surface", "salinity_depth", "o2_depth", "chl_surface", "mlp_surface",  "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
presback <- subset(presback, select = c(allvar))
colnames(presback)
```

@@

```{r remove na vals}
prback <- na.omit(presback)
```

@@

```{r train and test inc AIC for reg val}
aicetc <- read.csv("../data/env/AICcCalculations.csv", header = TRUE) #pre-created csv file
betavallist <- cbind(aicetc$regularization)
timeslice <- "all_data"
train <- prback
trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling
nfolds <- cbind(c(1:5))

  for (h in 1:nrow(betavallist)){
    betamval <-  betavallist[h]
    mod <- maxent(x = train, p = trainpa, path = paste0("../output/maxent/", timeslice, "/models"), args=c("maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", paste0("replicates=", nrow(nfolds)), "replicatetype=crossvalidate", paste0("betamultiplier=", betamval)))

@@

for(f in 1:nrow(nfolds)){
  modn <- nfolds[f]-1
      lamf <- lambdas_list[f]
      sampf <- samppred_list[f]
      coefile <- read.csv(paste0(lamf), header = FALSE, stringsAsFactors=FALSE)
      coefile <- head(coefile, -4)
      coefile2 <- subset(coefile, V2 != 0)
      nocoeff <- nrow(coefile2)
      bick <- sum(as.numeric(coefile2[, 2])) #dodgy 
      bic <- (bick*log(sum(trainpa == 1)) - 2*11) #dodgy
      #natural log
      samplepred <- read.csv(paste0(sampf), header = TRUE)
      samplepred$nat_log <- log(samplepred$Raw.prediction)
      sumsamplepred <- sum(samplepred$nat_log)

@@

@@

      aicetc$train_AUC <- trainauc
      aicetc$test_AUC <- testauc
      aicetc$n <- notrainingpoints
      aicetc$coefficients <- nocoeff
      aicetc$sum.of.logs <- sumsamplepred

      aicetc$bic <- bic #not sure this is correct...
      aicetc$beta <- betamval
      aicetc$foldno <- modn

    #calculate minimal predicted area using ecospat
    aicetc$mpa_0.9_cum <- ecospat.mpa(samplepred$Cumulative.prediction, perc = 0.9) #90%

    aicetc$AIC <- ((2*aicetc$coefficients)-(2*aicetc$sum.of.logs))
    aicetc$AICc <- (-2*(aicetc$sum.of.logs)+((2*aicetc$coefficients)*(aicetc$n/(aicetc$n - aicetc$coefficients-1))))
    write.csv(aicetc, paste0("../output/maxent/", timeslice, "/models/", betamval, "_", modn, "_", "AICforMaxentReg.csv"), row.names = FALSE)
      }

  # aa <- paste0("fold", modn, "_", betamval, "_", "AICforMaxentReg.csv")
    aicperbetalst <- list.files(paste0("../output/maxent/", timeslice, "/models/"), pattern = paste0("^", betamval, ".*AICforMaxentReg.csv"), full.names = TRUE)
    aicperbeta <- lapply(aicperbetalst, read.csv, header = TRUE)
    aicperbeta <- do.call(rbind, aicperbeta)
    aicbetamean <- colMeans(aicperbeta)
    aicperbeta <- rbind(aicperbeta, aicbetamean)
    aicperbeta$foldno[f+1] <- f+1
    write.csv(aicperbeta, paste0("../output/maxent/", timeslice, "/models/aicperbetasum_", betamval, ".csv"), row.names = FALSE)

    betaavglst <- list.files(paste0("../output/maxent/", timeslice, "/models/"), pattern = "aicperbetasum_.*csv", full.names = TRUE)
    bavg <- lapply(betaavglst, read.csv, header = TRUE)
    bavg <- do.call(rbind, bavg)
    bavg <- subset(bavg, foldno == f+1)
    write.csv(bavg, paste0("../output/maxent/", timeslice, "/models/avgbetalst.csv"), row.names = FALSE)
  }

>>>>

```{r remove unnecessary columns}
prback <- subset(presback, select = -c(cell_id, id, decimalLatitude, decimalLongitude, datecollected, institutioncode, individualcount, depth, resname, originalscientificname, collectioncode, day, nafo_zone, gear, longitude_meters, latitude_meters, total_cell_obs_xy, total_cell_obs_xyt, bottom_depth, XXtotal_cell_obs_xyzt, temp_celsius_depth, temp_celsius_surface, longitude_meters.1, latitude_meters.1, bottom_depth_glorys, longitude_meters.2, latitude_meters.2, cell_id_3d, total_cell_obs_xyzt, cell_id_xyzt, depthlayerno, depth_layer))
colnames(presback)
```

@@

```{r}
allvar <- c("occurrence", "year", "month", "temp_surface", "temp_depth", "salinity_surface", "salinity_depth", "o2_depth", "chl_surface", "mlp_surface",  "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
prback <- subset(prback, select = c(allvar))
colnames(prback)
```

@@

```{r remove na vals}
prback <- na.omit(prback)

prback <- subset(prback, month == 8)

presyrs <-  subset(prback, occurrence == 1) #this is to get years with complete presence data 
presyrs <- unique(presyrs$year) #make a vector of unique years
prback <- prback[prback$year %in% presyrs, ] #keep only the data for which there is precence years
```

@@

```{r train and test inc AIC for reg val}
aicetc <- read.csv("../data/env/AICcCalculations.csv", header = TRUE) #pre-created csv file
#betavallist <- cbind(aicetc$regularization)
betavallist <- seq(0.25, 0.75, by = 0.25)

timeslice <- "all_data"
train <- prback
trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling
nfolds <- cbind(c(1:2))

 for (h in 1:length(betavallist)){
   betamval <-  betavallist[h]
    #betamval <-  0.5
    mod <- maxent(x = train, p = trainpa, plots = FALSE, path = paste0("../output/maxent/", timeslice, "/models"), args=c("maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", paste0("replicates=", nrow(nfolds)), "replicatetype=crossvalidate", paste0("betamultiplier=", betamval)))

@@

@@

    maxresults <- read.csv(paste0("../output/maxent/", timeslice, "/models/maxentResults.csv"), header = TRUE)

for (i in 1:nrow(aicetc)){   
for(f in 1:nrow(nfolds)){
    modn <- nfolds[f]-1
      lamf <- lambdas_list[f]
      sampf <- samppred_list[f]
      coefile <- read.csv(paste0(lamf), header = FALSE, stringsAsFactors=FALSE)
      coefile <- head(coefile, -4)
      coefile2 <- subset(coefile, V2 != 0)
      nocoeff <- nrow(coefile2)
      bick <- sum(as.numeric(coefile2[, 2])) #dodgy 
      bic <- (bick*log(sum(trainpa == 1)) - 2*11) #dodgy
      #natural log
      samplepred <- read.csv(paste0(sampf), header = TRUE)
      samplepred$nat_log <- log(samplepred$Raw.prediction)
      sumsamplepred <- sum(samplepred$nat_log)

@@

   if(aicetc$nfold[i] == modn){
      aicetc$Train_AUC[i] <- trainauc
      aicetc$Test_AUC[i] <- testauc
      aicetc$n[i] <- notrainingpoints
      aicetc$coefficients[i] <- nocoeff
      aicetc$sum.of.logs[i] <- sumsamplepred

      aicetc$bic[i] <- bic #not sure this is correct...
      aicetc$beta[i] <- betamval
      aicetc$foldno[i] <- modn

    #calculate minimal predicted area using ecospat
    aicetc$mpa_0.9_cum[i] <- ecospat.mpa(samplepred$Cumulative.prediction, perc = 0.9) #90%

    aicetc$AIC[i] <- ((2*aicetc$coefficients[i])-(2*aicetc$sum.of.logs[i]))
    aicetc$AICc[i] <- (-2*(aicetc$sum.of.logs[i])+((2*aicetc$coefficients[i])*(aicetc$n[i]/(aicetc$n[i] - aicetc$coefficients[i]-1))))
   }

  }
    write.csv(aicetc, paste0("../output/maxent/", timeslice, "/models/", "_", betamval, "_AICforMaxentReg.csv"), row.names = FALSE)

}
  }

    aicperbetalst <- list.files(paste0("../output/maxent/", timeslice, "/models/"), pattern = paste0(".*AICforMaxentReg.csv"), full.names = TRUE)
    for (beta in 1:length(aicperbetalst)){
      aic <- read.csv(aicperbetalst[beta])
      bval <- str_sub(aicperbetalst[beta], 35, -20) #the year being used for testing
      aicmean <- colwise(mean)(aic)
      write.csv(aicmean, paste0("../output/maxent/", timeslice, "/models/aic_mean_", bval, ".csv"), row.names = FALSE)
    }

    aicmeanlst <- list.files(paste0("../output/maxent/", timeslice, "/models/"), pattern = paste0("^aic_mean"), full.names = TRUE)
    bavg <- lapply(aicmeanlst, read.csv, header = TRUE)
    bavg <- do.call(rbind, bavg)
    write.csv(bavg, paste0("../output/maxent/", timeslice, "/models/avgbetalst.csv"), row.names = FALSE)
